---
title: "Information Theory Responses"
output:
  html_document:
    df_print: paged
---

# Post-Workshop Survey

We conducted an online survey after the workshop, we had 8 respondants of ~40 participants (~20% response rate), and we wanted to know

1) was the workshop clear and useful?
2) would they recommend it to others?
3) what did they like and what can be improved?

```{r}
library(tidyverse)
data=read_csv("data/responses.csv") %>%
  rename(useful=`Overall, how useful was the information theory workshop?`,
         organization=`Please rate the overall organization of the material`,
         clarity=`Please rate the overall clarity of the material`,
         recommend=`Would you recommend this workshop to others?`,
         liked=`What were things that you liked about the workshop?`,
         improve=`What are ways that we can improve the workshop?`) %>%
  mutate(date=lubridate::mdy_hms(Timestamp)) 

```

## Usefulness

`Overall, how useful was the information theory workshop? (scale 1-10)`

```{r}
ggplot(data, aes(useful))+
  geom_density(size=2)+
  theme_minimal() +
  xlim(0, 10)
```

Looks like most people found it somewhat useful.

## Organization

`Please rate the overall organization of the material (scale 1-10)`

```{r}
ggplot(data, aes(organization))+
  geom_density(size=2)+
  theme_minimal() +
  xlim(0, 10)
```

Some people weren't fond of the organization, but overall people seemed to like it.

## Clarity

`Please rate the overall clarity of the material (scale 1-10)`

```{r}
ggplot(data, aes(clarity))+
  geom_density(size=2)+
  theme_minimal() +
  xlim(0, 10)
```

Again, some people thought it wasn't very clear, but others were fairly happy.

## Recommend?

`Would you recommend this workshop to others? (No/Maybe/Yes)`

```{r}
ggplot(data, aes(recommend)) +
  stat_count() +
  theme_minimal()
```

Most people would recommend it to others, but some are hesitant.

## Positives

`What were things that you liked about the workshop? (free response)`

```{r}
print(data$liked)
```

Some people weren't fond of my section (I tend to be a bit frantic during meetings, blame it on lack of sleep), but overall people seemed to like it.

## To Improve

`What are ways that we can improve the workshop?`

```{r}
print(data$improve)
```

We probably should have made the material available sooner, that was my failing, I was still learning how to do websites with Github.

## Conclusion

Overall, it seems like the workshop was well received, and that people found it useful. Now that the materials are prepared, it would be fairly straightforward to do another workshop in a similar vein.

## R Session Information

```{r}
sessionInfo()
```