---
title: Spinach Downy Mildew Case Study and Summary Statistic Comparisons
author: "Gareth Hughes, Robin Choudhury, Neil McRoberts"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    df_print: paged
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE) #change this to FALSE to suppress printing the code
library(tidyverse)
library(broom)
library(pROC)
```

Calculation of pseudo Rsq statistics from logistic regression. Analyses associated with ICPP2018 workhshop on Information Theory and a Letter to the Editor of Phytopathology. Analyses here based on data and original work from two sources:  
1.Choudhury et al. (2016). Logistic regression for disease incidence increase subsequent to catches of *Peronospora effusa* DNA detected on rotating arm spore traps.
2. De Wolf et al. (2003), also used in Madden (2006) and McRoberts et al, (2011).  Early data set used in constructing predictors of fusarium head blight (FHB) in wheat.


For the two datasets the same set of analyses will be carried out. First, a logistic regression will be performed with a binary disease state as the dependent variable and a single continuous variable as the predictor.  In the first example, the predictor is the natural logarithm of the estimated pathogen DNA copy number measured on spore traps prior to observations of disease state.  In the second and third examples the explanatory variable is a composite "risk" indicator constructed from a number of weather variables which individually correlate with disease incidence.  In the third example, the dependent binary disease variable was constructed by collapsing the original 5-point (0 to 4) ordinal disease intensity variable to two categories by combining the original 0 and 1 classes into a "low" disease class (designated 0) and the remaning classes, 2,3 and 4, into a "high" disease class (designated 1).


Diagnostic plots for each analysis were produced showing (1) the binary disase state and fitted probability values together against the explanatory variable, and (2) a histogram of the fitted probability values classified by the binary disease state and binned according to system defaults.


The analyses were run on R 3.3.3 for Windows under R Studio running on a Windows virtual machine on a remote application server accessed via Parallels Client for Chrome OS [ref for Parallels].  


```{r}
# install.packages("DescTools") Remove hashtag to run if needed, only need to run this once
# library("DescTools") #nice, wish I had seen this sooner rac

# Analysis 1 SDM data from Choudhury et al 2016.
# Get the data and take a look at the top of the data file

data<-read_csv("data/roc.test.csv") #read.csv was throwing out some weird errors for me, so im using read_csv from the readr package contained within tidyverse
attach(data)
head(data)

# Fit the basic logistic regression on Disease using Spore.4

dismodel1 <- glm(Disease~`Spore-4`,family=binomial(link='logit'), data = data) #I needed to put tics (`) around Spore-4 because when i loaded it with read_csv it read it as 'Spore-4', not 'Spore.4' rac
summary(dismodel1)
tidy(dismodel1) #rac tidy() and glance() are nice functions when you want to look at cleaner output
```


The Spore trap DNA count lagged by 4 observation periods (roughly 8-12 days) does have some positive association with subsequent disease state.  Although the estimated effect of lagged spore trap signal is unlikely to be due to chance alone the odds of disease do not increase strongly with increase in the signal.  


###Logistic Regression



```{r}
# Take a look at the fitted probability values from the regression.
fits1<-as.numeric(dismodel1$fitted.values) # need this because fitted.values is a list and while hist understands what
                                           # to do with it if you don't monkey about with it, asking it to sub-divide
                                           # the vector of values using the logical == confuses it, so we have to 
                                           # pull the numerical bit out of the list.
dis_state<-c(rep(1,46),rep(0,37))
dis_state2=dismodel1$model$Disease #this pulls it directly from the model rac
spore4_rac=dismodel1$model$`Spore-4` #this pulls it directly from the model rac

#I'm going to redo this in ggplot/tidyr feel free to ignore RAC
SDM_rac=data.frame(fits1=fits1, disease=dis_state2, spore4=spore4_rac)
SDM_rac1=SDM_rac %>%
  mutate(.,status=ifelse(disease==0, "Non-cases", "Cases"))

SDM_rac1$status <- factor(SDM_rac1$status, levels = c("Cases","Non-cases"))

ggplot(SDM_rac1, aes(fits1, fill=status))+
 geom_histogram(color="black")+
  scale_x_continuous(limits=c(0,1))+
  facet_grid(status~., scales = "free_y")+ 
  xlab("Fitted Probability Value")+
  scale_fill_manual(values = c("grey20", "grey80"))+
    theme_bw()+
  theme(legend.position = "none",
        legend.background = element_blank(),
        strip.text = element_text(face="bold", size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        plot.title = element_text(size = 16, face = "bold"))+
  guides(fill=guide_legend(title = "Status"))
 
#ggsave(filename = "figures/sdm_dot_hist_facet.png")


  ggplot(SDM_rac1, aes(fits1, fill=status))+
  geom_histogram(alpha=0.4, position="identity")+
  scale_x_continuous(limits=c(0,1))+
  #facet_grid(~status)+ 
  xlab("Probability")+
  ylab("Counts")+
  ggtitle("Histogram of Fitted Probabilities, SDM Data")+
  theme(legend.position = c(.15,.8),
        legend.background = element_blank(),
        strip.text = element_text(face="bold", size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        plot.title = element_text(size = 16, face = "bold"))+
  guides(fill=guide_legend(title = "Status"))
  
 # ggsave(filename = "figures/sdm_hist_identity.png")
```

```{r}
  # Generate a plot of binary disease variable and fitted probabilities
# There are four NA values corresponding to zeros in the data.  Dump
# the NA values and make a reduced vector for plotting the results
#Spore4<-na.omit(data$`Spore-4`) #couldnt get this to work rac
#plot(Spore4,dismodel1$fitted.values, xlab="ln(DNA copies)", ylab="Probability", col="dark blue",ylim=c(0,1))
#points(Spore.4,Disease, col="red")

SDM_rac_spore_melt=SDM_rac %>%
  gather(est.type, disease, -spore4)

ggplot(SDM_rac_spore_melt, aes(spore4, disease, color=est.type))+
  geom_point()+
  xlab("Log of Spore Copy Number")+
  ylab("Probability")+ #not sure this is the right label? rac
  ggtitle("Fitted vs. Real Disease, SDM Data")+
  theme(legend.position = c(.2,.8),
        legend.background = element_blank(),
        strip.text = element_text(face="bold", size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        plot.title = element_text(size = 16, face = "bold"))+
  guides(color=guide_legend(title = "Estimate Type"))+
  scale_color_discrete(labels=c("Disease Incidence","Fitted Disease Data"))
  #ggsave(filename = "figures/sdm_real_fit.png")


```

```{r}
#build an ROC for the SDM data
sdm_data<- read_csv("data/roc.test.csv")
sdm_roc=roc(sdm_data$Disease, sdm_data$Spore-4)
summary(sdm_roc)
plot(sdm_roc)
```

The difference between pROC and other software (e.g. - ROCR) can lead to differences between output ROCs.


```{r}
sessionInfo()
```


